wordmaker

### 2020-07-20 apa_style.Rmd
pre results edits

---
title: "An interdisciplinary examination of stress and injury occurrence in athletes"
runtitle: "Short Title"
documentclass: frontiersHLTH  # or frontiersHLTH, or frontiersFPHY
author:
  - name: Dr Harry Fisher
    affiliation: '1'
    etal: Fisher # First author's last name. 
    email: harryfisher21@gmail.com # Indicates corresponding Author
    institution: Cardiff Metropolitan University
    street: Cyncoed Road
    zip: CF23 6XD
    country: United Kingdom
  - name: Dr Marianne Gittoes
    affiliation: '1'
  - name: Professor Lynne Evans
    affiliation: '1'
  - name: Miss Leah Bitchell
    affiliation: '1'  
  - name: Dr Richard Mullen  
    affiliation: '2'
  - name: Dr Marco Scutari
    affiliation: '3'    
affiliation:
  - id: '1'
    department: School of Sport and Health Science
    institution: Cardiff Metropolitan University
    city: Cardiff
    country: United Kingdom
  - id: '2' 
    department: Division of Sport, Health & Exercise Sciences
    institution: Brunel University
    city: London
    country: United Kingdom
  - id: '3' 
    institution: Istituto Dalle Molle di Studi sull'Intelligenza Artificiale (IDSIA) 
    city: Manno
    country: Switzerland

output: rticles::frontiers_article
csl: frontiers.csl
bibliography: library.bib
---

\begin{abstract}

This paper adopts a novel, interdisciplinary approach to explore the relationship between psychosocial factors, physiological stress-related markers and occurrence of injury in athletes using a repeated measures design across a 2-year data collection period. At three data collection time-points, athletes completed measures of major life events, the reinforcement sensitivity theory personality questionnaire, muscle stiffness, heart rate variability and postural stability, an
d reported any injuries they had sustained since the last data collection. Two Bayesian networks were used to examine the relationships between variables and model the changes between data collection points in the study. Findings revealed muscle stiffness to have the strongest relationship with injury occurrence, with high levels of stiffness increasing the probability of sustaining an injury. Negative life events did not increase the probability of injury occurrence at any single time-point; however, when examining changes between time points, increases in negative life events did increase the probability of injury. In addition, the combination of increases in negative life events and muscle stiffness resulted in the greatest probability of sustaining an injury. Findings demonstrated the importance of both an interdisciplinary approach and a repeated measures design to furthering our understanding of the relationship between stress-related  markers and injury occurrence.

\tiny
 \keyFont{ \section{Keywords:} Sports Injury, Stress, Interdisciplinary, Bayesian Network}

\end{abstract}

```{r include=FALSE}
options(scipen=999, digits = 2)
knitr::opts_chunk$set(echo = FALSE)
# source("r-files/functions.R")
# source("r-files/methods_code.R")
# source("r-files/results_code.R")
load("all_data.RData")
pacman::p_load(tidyverse, knitr, kableExtra, qwraps2, brms, broom.mixed, bnlearn, tidyselect, Rgraphviz)
library(readxl)
library(janitor)
library(irr)
library(lavaan)
library(caret)
library(parallel)
library(zoo)
library(sjmisc)

@@

Over the last four decades sport related injuries have received increased research attention [@Ivarsson2017].
This attention is unsurprising given the high incidence [@Rosa2014; @Sheu2016],
and undesirable physical and psychological effects of sports injuries [@Leddy1994; @Brewer2012].
To mitigate against both the increasing incidence and	undesirable consequences of injury, research  has identified several psychological [@Slimani2018],
anatomical [@Murphy2003],
biomechanical [@Neely1998; @Hughes2014]
and environmental [@Meeuwisse2007]
factors associated with sports injury occurrence.
Indeed, several models of injury causation have been proposed that highlight the multifactorial nature of injury occurrence [@Kumar2001; @Meeuwisse2007; @Wiese-Bjornstal2009],
of which one of the most widely cited was developed by Williams and Anderson [Fig \ref{fig:fig1}; @Andersen1988; @Williams1998].

Williams and Andersen's [@Williams1998] model proposed that when faced with a potentially stressful athletic situation, an athlete’s personality traits (e.g., hardiness, locus of control and competitive trait anxiety), history of stressors (e.g., major life events and previous injuries) and coping resources (e.g., general coping behaviours) will contribute to their response, either interactively or in isolation.
Central to the model is the stress response, which reflects the bi-directional relationship between athletes’ appraisal of, and response to, a stressful athletic situation. The model predicts that athletes who have a history of many stressors, personality traits that intensify the stress response and few coping resources, will exhibit greater attentional (e.g., peripheral narrowing) and/or physiological (e.g., increased muscle tension) responses that put these individuals at greater risk of injury.

Within Williams and Andersen’s [@Williams1998] model, major life events, a component of an athlete’s history of stressors, most consistently predicts injury occurrence [@Williams2007];
specifically, major life events with a negative, as opposed to positive, valence [@Passer1983a; @Maddison2005].
However, personality traits and coping resources have also been	found to predict injury, with for example, athletes more likely to sustain an injury if they have poor social support and psychological coping skills, and high trait anxiety and elevated competitive state anxiety; compared to athletes with the opposite profile. [@Smith1990; @Lavallee1996; @Ivarsson2010].
However, the amount of variance explained by the psychosocial factors proposed by the model has been modest, typically between 5 - 30% [@Galambos2005; @Ivarsson2010]; suggesting other factors are also likely to contribute to injury occurrence.

While the psychosocial factors proposed in Williams and Andersen’s [@Williams1998] model have received the most research attention, the mechanisms through which these factors are proposed to exert their effect have remained under-investigated in the literature. 
To elaborate, the  model suggests that injuries are likely to occur through either increased physiological arousal resulting in increased muscle tension and reduced flexibility or attentional deficits caused by increased distractibility and peripheral narrowing.
However, to date, the research has largely focused on attentional deficits [@Andersen1999; @Rogers2005; @Wilkerson2012a; @Swanik2007].
For example, Andersen and Williams [@Andersen1999]
measured peripheral and central vision during high and low stress conditions and found athletes with high life event stress coupled with low social support had greater peripheral narrowing under stressful conditions compared to athletes with the opposing profile; these athletes went on to sustain an increased number of injuries during the following season.
Indeed, Rodgers and Landers [@Rogers2005] supported Andersen and Williams's [@Andersen1999] earlier findings reporting that peripheral narrowing under stress mediated 8.1% of the relationship between negative life events and injury.
However, the remaining variance between negative life events and athletic injury through the other proposed mechanisms, such as increased muscle tension and reduced motor control, remains to be explored [cf. @Williams1998].

One challenge faced by researchers addressing the sports injury problem with a psychological lens is the multifactorial nature of injury, and the possible contribution of other non-psychological factors to the stress response
[@Meeuwisse2007; @Wiese-Bjornstal2009].
For example, a large body of research indicates that training-related stress is also likely to be related to the stress response and injury occurrence [@Lee2017; @Djaoui2017],
and may account for the unexplained variance from the psychological predictors of injury.
Appaneal and Perna [@Appaneal2014]
proposed the biopsychosocial model of stress athletic injury and health (BMSAIH) to serve as an extension to Williams and Andersen’s [@Williams1998] model and to address some of these issues.
To elaborate, the BMSAIH aimed to clarify the mediating pathways between the stress response and injury, consider other health outcomes and behavioural factors that impact sports participation,
and integrate the impact of training on athletes’ health [@Appaneal2014].
The central tenet of the BMSAIH is that psychosocial distress (e.g., negative life events) may act synergistically with training-related stress as a result of high-intensity and high-volume sports training, and “widen the window of susceptibility” [@Appaneal2014, p. 74] to a range of undesirable health outcomes including illness and injury.
Consequently, the BMSAIH provides a framework for future research to build on Williams and Andersen's [@Williams1998] model, by including other physiological markers of training-related stress, which together may provide greater insight into the injury process.

Although research supporting the BMSAIH has mainly focused on the relationship between hormonal responses to training and injury occurrence [@Perna1995; @Perna1997; @Perna2003], other research has identified additional markers of training-related stress that are associated with an increased risk of injury; for example,
heart rate variability [@Bellenger2016; @Williams2017],
postural stability [@Romero-Franco2014]
and muscle stiffness [@Pruyn2015].
However, these markers are often studied in isolation without an assessment of the psychosocial factors that are known to contribute to injury, thereby limiting our understanding of how psychosocially and physiologically derived stress may contribute synergistically to injury occurrence.
Recently, Bittencourt et al. [@Bittencourt2016]
suggested that to better understand the multifactorial nature of sports injuries, research needs to move away from studying risk factors in isolation and instead adopt a complex systems approach to injury.
Such an approach posits that injury may arise from a complex “web of determinants” [@Bittencourt2016, p. 3],
where different factors interact in unpredictable and unplanned ways, but result in a global outcome pattern of either adaptation or injury.

A challenge when adopting a complex systems approach is using an appropriate analysis technique that is able to capture the uncertainty and complexity of the relationships between different variables. One technique that provides a solution to this challenge is Bayesian network (BN) modelling. BN's allow the construction of graphical probabilistic models using the underlying structure that connects different variables [@Scutari2014]. The learned structure can be used for inference by obtaining the posterior probabilities of a particular variable for a given query (e.g., if the value of variable A is x and the value of variable B is y, what is the probability variable C being value z?). Furthermore, BN's do not distinguish between dependent and independent variables as they are a form of unsupervised learning, which is a strength over regression or structure equation models when the underlying relationship in the network may not be known [@Olmedilla2018].

To summarise, despite offering a possible framework to build on the research stemming from Williams and Andersen's [@Williams1998] model, there remains and opportunity to explore other physiological stress-related markers proposed by the BMSAIH, in addition to the already well-established psychological characteristics known to be related to injury [@Appaneal2014].
Furthermore, research has typically not captured changes in both psychosocial factors and stress-related physiological markers that may occur between initial measurement and injury occurrence.
Given the exploratory nature of the current study the following objectives were defined:

- Identify suitable markers of stress that can be easily captured in a large cohort of athletes in a field based setting. 
- Capture the markers of stress and injury occurrence using a prospective, repeated measures design.
- Explore and evaluate the relationships between the markers of stress and injury using Bayesian network modelling.

@@

A total of 351 athletes (male: _n_ = 231, female: _n_ = 120) were initially recruited for the study. Participants had an average age of `r mean_sd(parchar$age, digits = 1)` years and represented a range of team (football, rugby, netball, cricket, lacrosse, basketball and field hockey) and individual sports (athletics, tennis, weightlifting, gymnastics, judo, swimming and golf from a British University and local sports clubs (Table \ref{tab:table1}).
Participants self-rated competitive level ranged from recreational to international standard. 
A total of `r n_perc(parchar$pic == "1")` participants had sustained an injury in the 12 months prior to the start of the study (male: _n_ = 114 [49%], female; _n_ = 48 [40%]).
At the start of the study, all participants were injury free (no modifications to their usual training routine due to a sport related medical problem for a minimum of four weeks).
Participants were engaged in training for their respective sports for at least five hours per week.
Ethical approval was obtained from the University ethics committee prior to the start of the study and all participants provided informed consent.

@@

A modified version of the Life Events Survey for Collegiate Athletes (LESCA) was used to measure participants' history of life event stress [@Petrie1992]. 
Modifications were made to the LESCA to ensure the suitability of the items for the study population (S1 Table).
The LESCA comprises 69 items that reflect possible life events that participants may have experienced. 
Example items include, "Major change in the frequency (increased or decreased) of social activities due to participation in sport", "Major change in the amount (more or less) of academic activity (homework, class time, etc)" and "Major change in level of athletic performance in actual competition (better or worse)".
Participants were asked to rate the perceived impact of each life event they had experienced within the last 12 months on an 8-point Likert scale anchored at -4 ($extremely\;negative$) and +4 ($extremely\;positive$). 
Negative and positive life event scores were calculated by summing the negative and positive scores respectively. 
A score for total life events was calculated by summing the absolute values for both negative and positive events.
Petrie (1992) [@Petrie1992] reported test-retest reliabilities at 1-week and 8-weeks with values ranging from .76 to .84 (p < .001) and .48 to .72 (p < .001) respectively. 
Petrie also provided evidence of discriminant, convergent and predictive validity. 
The LESCA is the most widely used measure of major life events for athletes in the sports injury literature. 
For this study, Composite Reliability [@Fornell1981] was preferred to Cronbach’s alpha as it does not assume parallelity (i.e., all factor loadings are constrained to be equal, and all error variances are constrained to be equal) and instead takes into consideration the varying factor loadings of the items in the questionnaire.
The composite reliability for the LESCA in this study was 0.84.

@@

A revised version of the Reinforcement Sensitivity Theory Personality Questionnaire (RST-PQ) was used to measure motivation, emotion, personality and their relevance to psychopathology [@Corr2016c]. 
The revised version of the RST-PQ comprises 51 statements that measure three major systems: Fight-Flight-Freeze System (FFFS; e.g., "I am the sort of person who easily freezes-up when scared"), Behavioural Inhibition System (BIS; e.g., "When trying to make a decision, I find myself constantly chewing it over") and four Behavioural Approach System (BAS) factors; Reward Interest (e.g., “I regularly try new activities just to see if I enjoy them”), Goal Drive Persistence (e.g., “I am very persistent in achieving my goals”), Reward Reactivity (e.g., “I get a special thrill when I am praised for something I've done well”) and Impulsivity (e.g., “I find myself doing things on the spur of the moment”).
Participants rated each item on a scale from 1 ($not\; at\; all$) to 4 ($highly$) to reflect how well each statement described their personality in general. 
The responses to items associated with each subscale (FFFS, BIS, RI, GDP, RR and I) were summed to give a total score that was subsequently used for further analysis.
The composite reliabilities for each subscale were; BIS = `r rst_rel$final[1]`, FFFS = `r rst_rel$final[2]`, GDP = `r rst_rel$final[3]`, I = `r rst_rel$final[4]`, RI = `r rst_rel$final[5]`, RR = `r rst_rel$final[6]`. Further details regarding the revised RST are in S1 Appendix.

@@

A handheld myometer (MyotonPRO, Myoton AS, Tallinn, Estonia) was used to measure muscle stiffness.
The MyotonPRO is a non-invasive, handheld device that applies a mechanical impulse of 0.40 N for 0.15 ms perpendicular to the surface of the skin.
The impulse causes natural damped oscillations in the tissue, which are recorded by a three-axis digital accelerometer sensor in the device.
The raw oscillation signal is then processed, and the stiffness parameter is calculated [@Agyapong-Badu2016]. 
The MyotonPRO has previously been reported to be a reliable and valid tool for the measurement of in-vivo tissue stiffness properties [@Chuang2013; @Pruyn2016; @Nair2014], and has demonstrated good internal consistency (coefficient of variation < 1.4%) over sets of 10 repetitions [@Aird2012].

@@

The error identification criteria from the original BESS protocol was used by the lead researcher who scored all the BESS trials. 
One error was recorded if any of the following movements were observed during each trial: a) lifting hands off iliac crests; 
b) opening eyes; 
c) stepping, stumbling, or falling; 
d) moving the thigh into more than 30 degrees of flexion or abduction; 
e) lifting the forefoot or heel; and 
f) remaining out of the testing position for more than 5-seconds [@Riemann1999d].
A maximum score of 10 errors was possible for each stance. Multiple errors occurring simultaneously were recorded as one error.
A participant was given the maximum score of 10 if they remained out of the stance position for more than 5-seconds.
To calculate limb asymmetry, the DL and NDL leg score was calculated by summing the DL and NDL errors respectively.
A total score was calculated by summing the total number of errors recorded on all stances (DL, NLD and TS, on foam and firm surfaces). 
To assess the intra-rater reliability, a single measurement, absolute agreement, two-way mixed effects model for the intraclass correlation [ICC; @Koo2016] was used on a sample of 40 participants from the first time point.
The test-retest scoring of BESS resulted in a "good" to "excellent" ICC score (ICC = `r round(icc[["value"]], digits = 2)`, 95% confidence interval = `r paste(round(icc[["lbound"]], digits = 2), "-", round(icc[["ubound"]], digits = 2)) `), indicating the scoring was reliable.

@@

At the start of the UK academic year (September 2016 and 2017), coaches of sports teams at a British University and local sports clubs were contacted and informed about the study. 
With the coaches' permission, the lead researcher attended training sessions to inform the athletes about the overall purpose of the study and the requirements of participation.
To be eligible for the study athletes had to be injury free (no modifications to their usual training routine due to a sport related medical problem for a minimum of four weeks) and training a minimum of five hours per week. 
Athletes who met the criteria and volunteered to take part in the study were invited to attend scheduled testing sessions.
A repeated measures prospective cohort design was used to assess athletes' major life events, stress-related physiological markers and injury status over two consecutive twelve-month periods between September 2016 and September 2018. 
Each participant was asked to attend a total of 3 data collections over a twelve-month period, with each data collection separated by a four-month interval (Fig \ref{fig:fig3}). 
Participants provided informed consent before data collection commenced.

@@

The questionnaires, which included demographic information, the LESCA, RST-PQ (T1, T2, T3) and injury status (T2, T3, T4) were completed  on-line (SurveyMonkey Inc., USA, www.surveymonkey.com).
The instructions for the LESCA were modified at T2 and T3 so that participants reported major life events that had occurred since the previous testing session.
For injury reporting, participants were asked to record any injuries that they had sustained since the last data collection.
The data were downloaded from surverymonkey.com and imported into R [@RCoreTeam2019] for analysis purposes.

@@

To minimise potential distractions, participants were directed to a designated quiet area in the laboratory where IBI data were recorded.
Participants were instructed to turn off their mobile devices to avoid any interference with the Bluetooth sensor.
Each chest strap was dampened with water and adjusted so it fitted tightly but comfortably, as outlined by Polar's guidelines. 
Participants were seated and asked to remain as still as possible for the duration of the recording. 
No attempt was made to control participants respiratory frequency or tidal volume [@Denver2007]. 
Inter-beat interval (IBI) data was collected for 10-minutes at a sampling frequency of 1000 Hz.

@@

Prior to the final network structure presented in the results section, several structures with different combinations of variables were explred.
Networks were learned using a Tabu search algorithm [@Norvig2009] and BIC was used to compare different models. 
A higher BIC value indicates the structure of a DAG is a better fit to the observed data [@Scutari2014]. 
BIC values for each combination of variables of interest are reported as the combination of variables with the highest BIC value, followed by the relative scores of the other variables in the model.

>>>>

---
title             : "An interdisciplinary examination of stress and injury occurrence in athletes"
shorttitle        : "Stress and sports injury"

author: 
  - name          : "Harry Fisher"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       :  "Cyncoed Road, Cardiff, CF23 6XD"
    email         : "harryfisher21@gmail.com"
  - name: Dr Marianne Gittoes
    affiliation: '1'
  - name: Professor Lynne Evans
    affiliation: '1'
  - name: Miss Leah Bitchell
    affiliation: '1'  
  - name: Dr Richard Mullen  
    affiliation: '2'
  - name: Dr Marco Scutari
    affiliation: '3'

affiliation:
  - id: '1'
    department: School of Sport and Health Science
    institution: Cardiff Metropolitan University
    city: Cardiff
    country: United Kingdom
  - id: '2' 
    department: Division of Sport, Health & Exercise Sciences
    institution: Brunel University
    city: London
    country: United Kingdom
  - id: '3' 
    institution: Istituto Dalle Molle di Studi sull'Intelligenza Artificiale (IDSIA) 
    city: Manno
    country: Switzerland

abstract: |
 This paper adopts a novel, interdisciplinary approach to explore the relationship between psychosocial factors, physiological stress-related markers and occurrence of injury in athletes using a repeated measures design across a 2-year data collection period. At three data collection time-points, athletes completed measures of major life events, the reinforcement sensitivity theory personality questionnaire, muscle stiffness, heart rate variability and postural stability, and reported any injuries they had sustained since the last data collection. Two Bayesian networks were used to examine the relationships between variables and model the changes between data collection points in the study. Findings revealed muscle stiffness to have the strongest relationship with injury occurrence, with high levels of stiffness increasing the probability of sustaining an injury. Negative life events did not increase the probability of injury occurrence at any single time-point; however, when examining changes between time points, increases in negative life events did increase the probability of injury. In addition, the combination of increases in negative life events and muscle stiffness resulted in the greatest probability of sustaining an injury. Findings demonstrated the importance of both an interdisciplinary approach and a repeated measures design to furthering our understanding of the relationship between stress-related  markers and injury occurrence.

keywords          : "Sports Injury, Stress, Interdisciplinary, Bayesian Network"

bibliography      : [library.bib, r-references.bib]

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
csl: frontiers.csl
always_allow_html: true
---

```{r setup, include = FALSE}
library("papaja")
options(kableExtra.auto_format = FALSE)
```

```{r include=FALSE}
options(scipen=999, digits = 2)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
# source("r-files/functions.R")
# source("r-files/methods_code.R")
# source("r-files/results_code.R")
load("all_data.RData")
pacman::p_load(tidyverse, knitr, kableExtra, qwraps2, brms, broom.mixed, bnlearn, tidyselect, Rgraphviz)
library(readxl)
library(janitor)
library(irr)
library(lavaan)
library(caret)
library(parallel)
library(zoo)
library(sjmisc)

@@

Over the last four decades sport-related injuries have received increased research attention [@Ivarsson2017] in response to the high incidence [@Rosa2014; @Sheu2016] and associated undesirable physical and psychological effects [@Leddy1994; @Brewer2012]. Multiple psychological [@Slimani2018], anatomical [@Murphy2003], biomechanical [@Neely1998; @Hughes2014] and environmental [@Meeuwisse2007] factors have been associated with sports injury occurrence and several models of injury causation have been proposed that highlight the multifactorial nature of injury occurrence [@Kumar2001; @Meeuwisse2007; @Wiese-Bjornstal2009], of which one of the most widely cited was developed by Williams and Anderson [Fig \ref{fig:fig1}; @Andersen1988; @Williams1998].

Williams and Andersen’s [@Williams1998] stress-injury model proposed that when faced with a potentially stressful athletic situation, an athlete’s personality traits (e.g., hardiness, locus of control and competitive trait anxiety), history of stressors (e.g., major life events and previous injuries) and coping resources (e.g., general coping behaviours) contribute to the injury response, either interactively or in isolation. The stress response is central to the model and reflects the bi-directional relationship between athletes’ appraisal of, and response to, a stressful athletic situation. The model predicts that athletes who have a history of stressors, personality traits that intensify the stress response and few coping resources, will exhibit greater attentional (e.g., peripheral narrowing) and/or physiological (e.g., increased muscle tension) responses that put these individuals at greater injury risk.

Within Williams and Andersen’s [@Williams1998] model, major life events, which is a component of an athlete’s history of stressors, most consistently predicts injury occurrence [@Williams2007]; specifically, major life events with a negative, as opposed to positive, valence [@Passer1983a; @Maddison2005]. Personality traits and coping resources have also been found to predict injury with athletes more likely to sustain an injury if they have poor social support and psychological coping skills, and high trait anxiety and elevated competitive state anxiety; compared to those with the opposing profile [@Smith1990; @Lavallee1996; @Ivarsson2010].However, the amount of variance explained by these psychosocial factors has been modest and typically between 5 - 30% [@Galambos2005; @Ivarsson2010], which indicates a likely interaction with other factors.

While the psychosocial factors in Williams and Andersen’s [@Williams1998] model have received the most research attention, less insight into the mechanisms through which these factors are proposed to exert their effect exists. To elaborate, the model suggests that injuries are likely to occur through either increased physiological arousal resulting in increased muscle tension and reduced flexibility or attentional deficits caused by increased distractibility and peripheral narrowing. However, to date, the research has largely focused on attentional deficits [@Andersen1999; @Rogers2005; @Wilkerson2012a; @Swanik2007]. For example, Andersen and Williams [@Andersen1999] found athletes with high life event stress coupled with low social support had greater peripheral narrowing under stressful conditions compared to athletes with the opposing profile; these athletes went on to sustain an increased number of injuries during the following athletic season. Rodgers and Landers [@Rogers2005] further supported Andersen and Williams’s [@Andersen1999] earlier findings by identifying that peripheral narrowing under stress mediated 8.1% of the relationship between negative life events and injury.

Knowledge of the physiological factors (e.g., increased muscle tension and reduced motor control) contributing to the remaining variance between negative life events and athletic injury remains sparse [cf. @Williams1998]. One challenge faced by researchers addressing the sports injury problem through a psychological lens is the multifactorial nature of injury, and the possible interaction with physiological factors in the stress response [@Meeuwisse2007; @Wiese-Bjornstal2009]. For example, a large body of research has suggested that training-related stress is also likely to contribute to the stress response and injury occurrence [@Lee2017; @Djaoui2017] and may account for the unexplained variance from the psychological predictors. Appaneal and Perna [@Appaneal2014] proposed the biopsychosocial model of stress athletic injury and health (BMSAIH) to serve as an extension to Williams and Andersen’s [@Williams1998] model. The BMSAIH enhances our understanding of the mediating pathways between the stress response and injury alongside other health outcomes and behavioral factors that impact sports participation [@Appaneal2014]. The central tenet of the BMSAIH is that psychosocial distress (e.g., negative life events) may act synergistically with training-related stress as a result of high-intensity and high-volume sports training, and “widen the window of susceptibility” [@Appaneal2014, p. 74] to a range of undesirable health outcomes including illness and injury. Consequently, the BMSAIH provides an important framework that has enhanced insight into the multi-faceted nature of the injury process by building on Williams and Andersen’s [@Williams1998] model whilst including other physiological markers of training-related stress.

Although research supporting the BMSAIH has mainly focused on the relationship between hormonal responses to training and injury occurrence [@Perna1995; @Perna1997; @Perna2003], other training-related stress markers including heart rate variability [@Bellenger2016; @Williams2017],  muscle stiffness [@Pruyn2015] and postural stability [@Romero-Franco2014] have been reported to be associated with an increased risk of injury. However, the largely mono-disciplinary approach to examining the role of these markers has provided limited insight into the multifaceted interactions with psychological factors that may contribute to injury occurrence. Furthermore, a reliance on designs that capture a single point of measurement precludes the assessment of intra as well as inter-individual changes and the effect of the time interval between measurement and injury occurrence on subsequent injuries [@Johnson2014]. Such an approach fails to capture changes in both psychological factors and physiological markers that may occur preceding an injury. Importantly, a repeated measurement approach would enable an assessing of how variables and their interactions change with respect to time provide greater insight into the effect that repeated exposure to major life events and other stress-related factors has on injury occurrence.

Recently, Bittencourt et al. [@Bittencourt2016] advocated a move away from studying isolated risk factors and instead, adopt a complex systems approach in order to understand injury occurrence. Such an approach posits that injury may arise from a complex “web of determinants” [@Bittencourt2016, p. 3], where different factors interact in unpredictable and unplanned ways, but result in a global outcome pattern of either adaptation or injury. Capturing the uncertainty and complexity of the relationships between different variables using an appropriate interdisciplinary analysis within the framework of a complex systems approach is challenging. Bayesian network (BN) modelling provides one solution by allowing the construction of graphical probabilistic models using the underlying structure that connects different variables [@Scutari2014]. The learned BN structure can be used for inference by obtaining the posterior probabilities of a particular variable for a given query (e.g., if the value of variable A is x and the value of variable B is y, what is the probability variable C of being value z?). Furthermore, unlike regression or structural equation models, BN’s do not distinguish between dependent and independent variables when the underlying relationship in the network may not be known [@Olmedilla2018]. BN modelling subsequently provides a valuable but underused interdisciplinary approach to investigating the complex and unpredictable interactions of psychological and physiological factors implicated in the injury process.

Using the frameworks provided by Williams and Andersen’s [@Williams1998] stress injury model and Appaneal’s BMSAIH model, the aim of this interdisciplinary study was to develop new understanding of the multifaceted interactions of psychological and physiological stress-related factors with injury occurrence. A prospective, repeated measures design incorporating field-based physiological and psychological measures combined with a BN modelling analysis was used to address the study aim.

@@

A total of 351 athletes (male: _n_ = 231, female: _n_ = 120) were initially recruited for the study from a British University and local sports clubs (Table \ref{tab:table1}). Participants had an average age of `r mean_sd(parchar$age, digits = 1)` years and represented a range of team (football, rugby, netball, cricket, lacrosse, basketball and field hockey) and individual sports (athletics, tennis, weightlifting, gymnastics, judo, swimming and golf. 
Participants self-rated competitive level ranged from recreational to international standard. 
They were engaged in training for their respective sports for at least five hours per week.
A total of `r n_perc(parchar$pic == "1")` participants had sustained an injury in the 12 months prior to the start of the study (male: _n_ = 114 [49%], female; _n_ = 48 [40%]).
All participants were injury free (no modifications to their usual training routine due to a sport related medical problem for a minimum of four weeks) at the start of the study. 
Ethical approval was obtained from the University ethics committee prior to the start of the study and all participants provided informed consent.

@@

A modified version of the Life Events Survey for Collegiate Athletes (LESCA) was used to measure participants’ history of life event stress [@Petrie1992]. 
The LESCA is the most widely used measure of major life events for athletes in the sports injury literature. 
Modifications were made to the LESCA to ensure the suitability of the items for the study cohort (S1 Table). 
The LESCA comprises 69 items that reflect possible life events that participants may have experienced. Example items include, “Major change in the frequency (increased or decreased) of social activities due to participation in sport”, “Major change in the amount (more or less) of academic activity (homework, class time, etc)” and “Major change in level of athletic performance in actual competition (better or worse)”. Participants were asked to rate the perceived impact of each life event they had experienced within 12-months preceding the study onset on an 8-point Likert scale anchored at -4  ($extremely\;negative$) and +4 ($extremely\;positive$). 
Negative and positive life event scores were calculated by summing the negative and positive scores, respectively. A score for total life events was also calculated by summing the absolute values for both negative and positive events. 
Petrie (1992) [@Petrie1992] reported test-retest reliabilities at 1-week and 8-weeks with values ranging from .76 to .84 (p < .001) and .48 to .72 (p < .001) respectively. Petrie also provided evidence of discriminant, convergent and predictive validity. For this study, composite reliability [@Fornell1981] was preferred to Cronbach’s alpha as it does not assume parallelity (i.e., all factor loadings are constrained to be equal, and all error variances are constrained to be equal) and instead takes into consideration the varying factor loadings of the items in the questionnaire. The composite reliability for the LESCA in this study was .84.

@@

A revised version of the Reinforcement Sensitivity Theory Personality Questionnaire (RST-PQ) was used to measure motivation, emotion, personality and their relevance to psychopathology [@Corr2016c]. 
The revised version of the RST-PQ comprises 51 statements that measure three major systems: Fight-Flight-Freeze System (FFFS; e.g., "I am the sort of person who easily freezes-up when scared"), Behavioural Inhibition System (BIS; e.g., "When trying to make a decision, I find myself constantly chewing it over") and four Behavioural Approach System (BAS) factors; Reward Interest (e.g., “I regularly try new activities just to see if I enjoy them”), Goal Drive Persistence (e.g., “I am very persistent in achieving my goals”), Reward Reactivity (e.g., “I get a special thrill when I am praised for something I've done well”) and Impulsivity (e.g., “I find myself doing things on the spur of the moment”).
Participants rated each item on a scale from 1 ($not\; at\; all$) to 4 ($highly$) to reflect how well each statement described their personality in general. 
The responses to items associated with each subscale (FFFS, BIS, RI, GDP, RR and I) were summed to give a total personality score that was subsequently used for further analysis.
The composite reliabilities for each subscale were; BIS = `r rst_rel$final[1]`, FFFS = `r rst_rel$final[2]`, GDP = `r rst_rel$final[3]`, I = `r rst_rel$final[4]`, RI = `r rst_rel$final[5]`, RR = `r rst_rel$final[6]`. Further details regarding the revised RST are in S1 Appendix.

@@

A handheld myometer (MyotonPRO, Myoton AS, Tallinn, Estonia) was used to measure passive muscle stiffness.
The MyotonPRO is a non-invasive, handheld device that applies a mechanical impulse of 0.40 N for 0.15 ms perpendicular to the surface of the skin.
The impulse causes natural damped oscillations in the tissue, which are recorded by a three-axis digital accelerometer sensor in the device.
The raw oscillation signal is then processed, and the stiffness parameter is calculated [@Agyapong-Badu2016]. 
The MyotonPRO has previously been reported to be a reliable and valid tool for the measurement of in-vivo tissue stiffness properties [@Chuang2013; @Pruyn2016; @Nair2014], and has demonstrated good internal consistency (coefficient of variation < 1.4%) over sets of 10 repetitions [@Aird2012].

@@

The error identification criteria from the original BESS protocol was used by the lead researcher who scored all the BESS trials. 
One error was recorded if any of the following movements were observed during each trial: a) lifting hands off iliac crests; 
b) opening eyes; 
c) stepping, stumbling, or falling; 
d) moving the thigh into more than 30 degrees of flexion or abduction; 
e) lifting the forefoot or heel; and 
f) remaining out of the testing position for more than 5-seconds [@Riemann1999d].
A maximum score of 10 errors was possible for each stance. Multiple errors occurring simultaneously were recorded as one error.
A participant was given the maximum score of 10 if they remained out of the stance position for more than 5-seconds.
A total score was calculated by summing the total number of errors recorded on all stances (DL, NLD and TS, on foam and firm surfaces). 
To assess the intra-rater reliability, a single measurement, absolute agreement, two-way mixed effects model for the intraclass correlation [ICC; @Koo2016] was used on a sample of 40 participants from the first time point.
The test-retest scoring of BESS resulted in a "good" to "excellent" ICC score (ICC = `r round(icc[["value"]], digits = 2)`, 95% confidence interval = `r paste(round(icc[["lbound"]], digits = 2), "-", round(icc[["ubound"]], digits = 2)) `), indicating the scoring was reliable.

@@

At the start of the academic year (September), coaches of sports teams at a British University and local sports clubs were contacted and informed about the study. With the coaches’ permission, the lead researcher attended training sessions to inform athletes about the overall purpose of the study and the requirements of participation. Athletes who met the participation criteria and volunteered to take part in the study were invited to attend scheduled testing sessions. A repeated measures prospective cohort design was used to assess athletes’ major life events, stress-related physiological markers and injury status over two consecutive 12-month periods. Each participant was asked to attend a total of three data collections over a 12-month period, with each data collection separated by a four-month interval (Fig \ref{fig:fig3}). Participants provided informed consent before data collection commenced.

@@

The questionnaires, which included demographic information, the LESCA, RST-PQ (T1, T2, T3) and injury status (T2, T3, T4) were completed  on-line (SurveyMonkey Inc., USA, www.surveymonkey.com).
The instructions for the LESCA were modified at T2 and T3 so that participants reported major life events that had occurred since the previous testing session.
For injury reporting, participants were asked to record any injuries that they had sustained since the last data collection.
The data were downloaded from surverymonkey.com and imported into R [@RCoreTeam2019] for analysis.

@@

To minimise potential distractions, participants were directed to a designated quiet area in the laboratory where IBI data were recorded.
Participants were instructed to turn off their mobile devices to avoid any interference with the Bluetooth sensor.
Each chest strap was dampened with water and adjusted so it fitted tightly but comfortably, as outlined by Polar's guidelines. 
Participants were seated and asked to remain as still as possible for the duration of the recording. 
No attempt was made to control participants' respiratory frequency or tidal volume [@Denver2007]. 
Inter-beat interval (IBI) data was collected for 10-minutes at a sampling frequency of 1000 Hz.

@@

Prior to the final network structure presented in the results section, several structures with different combinations of variables were explored. Networks were learned using a Tabu search algorithm [@Norvig2009] and BIC was used to compare different models. A higher BIC value indicates the structure of a DAG is a better fit to the observed data [@Scutari2014]. BIC values for each combination of variables of interest are reported as the combination of variables with the highest BIC value, followed by the relative scores of the other variables in the model.
